\chapter{Introduction to the Formalization of Fux's Theory}\label{ch:Formalization}
The formalization of Fux is done in several steps:
\begin{enumerate}[wide]
    \item \textbf{Spot the right rules in the \textit{Gradus Ad Parnassum}.} Fux tended to explain certain rules of music so that they were easy to understand and use for the musicians of the time. This implies that sometimes several rules can be reduced to one. On the other hand, some of the rules of music are not written as such in the book because they are implicit. For example, it goes without saying that counterpoint belongs to a certain key and scale, but this is never explicitly written in the book. In order not to create misunderstanding, it was decided to write them explicitly and separately in the next sections.
    \item \textbf{Formalize the rules in natural language in a way that is easy to express as constraints.} Indeed, the \textit{Gradus Ad Parnassum} is a work dedicated to a 17th century audience. It is necessary to read it with a critical eye and to translate it into modern language. That is, to reduce several rules into one, or at times, some rules are expressed in inclusive terms, whereas it is easier for a mathematician or computer scientist to write them in an equivalent way with exclusive terms or vice versa. Examples will be given in section \ref{subsec:1SPEnglish}.
    \item \textbf{On the one hand, write the rules in discrete mathematics.} This is a crucial step in order to be able to use these rules precisely in other contexts and with other programming languages. This will also allow us to check whether solutions exist mathematically. Indeed, some rules may be contradictory and, consequently, no solution would be possible. It is important to keep in mind that some rules are written in a way that can be easily written with the Gecode tool.
    \item \textbf{On the other hand, write the rules in constraint programming language.} The final goal of this thesis is to have constraints fixed according to Fux's rules and to find the best possible solutions with Gecode.
\end{enumerate}

\input{Sections/sct_ArraysNotation.tex}
\input{Sections/sct_Definitions.tex}
\input{Sections/sct_GeneralRules.tex}
\section{Types of rules}
Three types of rules are distinguished in the next chapters:
\begin{itemize}
    \item \textbf{Harmonic rules}: harmonic rules concern the harmonic intervals between the different voices, i.e. the harmony created by the \cf and the counterpoint of the same measure. They are often the most important and the most numerous. These rules are noted by the letter \textbf{H}.
    \item \textbf{Melodic rules}: melodic rules refer to the melodic intervals of counterpoint or \cfcomma which usually correspond to the gap between two consecutive notes of the same voice. These rules are noted by the letter \textbf{M}.
    \item \textbf{Motion or Harmonic and Melodic rules}: these rules use both of the above types of intervals. They are more complex and often relate to specific motions. These rules are noted by the letter \textbf{P} for \textit{passage} since \textit{M} is already taken\footnote{This way of classifying is only intended to clarify the idea behind a rule. It remains quite abstract and subjective because some rules are classified as melodic while they also use harmonic constraints. In no case does Fux make a delimitation between his explanations in this way.}.
\end{itemize}
The notation of the rules is: \textbf{S.TX} where \textbf{S} is the species, \textbf{T} is the type of rule (H, M or P), and \textbf{X} is the number of the rule. For example, the sixth harmonic rule of the first species is written \textbf{1.H6}.