\chapter{Introduction and context of this work}
\pagenumbering{arabic}
This thesis is a formalisation of three-part counterpoint based on the writings and rules of Fux. Its aim is to provide a mathematical set of rules and a computer environment capable of translating Fux's teachings into formal logic, and capable of implementing these logical rules in a concrete way to produce Fux-style counterpoint.


This thesis will therefore be divided into several parts: we will first immerse ourselves in \gap, Fux's central work, from which we will meticulously extract the rules laid down by its author. We will briefly discuss these rules to make them unambiguous, and then translate them into formal logic, so that each rule Fux had in mind when writing his work is mathematically recorded. On this basis, we will create a computer implementation using constraint programming. We will then look at how this implementation finds results, discussing the search algorithm and heuristics used. We then discuss the cost techniques used to obtain the best possible results. Finally, we will analyse the musical compositions produced by the tool created.

It is very important to know that this thesis is based on T. Wafflard's thesis "FuxCP: a constraint programming based tool formalizing Fux's musical theory of counterpoint" \cite{wafflard2023} and article "A Constraint Formalization of Fux's Counterpoint" \cite{sprockeels2023constraint}. The present work takes up the concepts and definitions of T. Wafflard and could only be understood in its full depth by reading and fully understanding his works as well. A short summary is given in section \ref{section:thomas-in-a-nutshell}.



\section{A brief history of counterpoint: from the writings of Bach to algorithmic generation}
And before we get to the heart of the matter (which is the formalisation), let's take a look at Fux's theory of counterpoint, because that's the work that this work aims to formalise. Counterpoint is a compositional technique in which there are several musical lines (or voices) that are independent and distinct from one another but are balanced and sound beautiful \cite{CpSachs}. No voice is dominant over the others, and all are main voices, even though some may take a small precedence during part of the composition \cite{hess2016}.  


Counterpoint has been central to the work of many famous composers from different artistic movements, such as Bach in the Baroque, Mozart in the Classical and Beethoven in the Romantic \cite{kramer1987gradus}, has reached some modern music \cite{altozano2017contrapunto}, and has aroused interest over the centuries with the development of key texts on the subject, such as Schenker's Counterpoint \cite{schenker1906} or Jeppesen's Analysis \cite{jeppesen1960}. And while Bach was probably the master of counterpoint composition in his day \cite{yearsley2002}, the central and foundational work in the teaching of counterpoint belongs to another great Baroque composer: the Austrian Johann Joseph Fux and his treatise \gap. In it, this Baroque composer gives a detailed analysis of two-, three- and four-part counterpoint writing, all told as a discourse between a master and his pupil. \gaps was part of the species counterpoint movement, a way of conceiving counterpoint in five different types that could then be combined. It is on this work that this dissertation is based.


For Fux, but also for many other authors, species counterpoint are governed by many diverse rules, and it is these rules that interest us in the present work. The rules are based on old concepts that go back to older styles and have been discussed by many authors \cite{crocker1962}.  Those concepts include, for example, the notions of opposite motion and consonance (which in turn can be either perfect or imperfect). These concepts and their application to counterpoint are particularly interesting because they allow us to consider the composition of counterpoint both in a 'vertical' way, in which we consider the harmony of the notes played together, and in a 'horizontal' way, in which we consider the melodic development of each of the parts individually, which provides the independence of the counterpoints from each other and their melodic beauty.

This is what makes it interesting to analyse from a constraint programming point of view. We'll come back to this later, but for now let's concentrate on Fux's music theory.

\subsection{Fux's theory of counterpoint}
As we have just said, Fux uses a set of rules to create 'the right counterpoint'. These rules can be divided into three categories: melodic rules, harmonic rules and rules of motion. We will examine them here, bearing in mind that the formalisation of all the rules for two-voice counterpoint can be found in T. Wafflard's thesis, and that the complete set of rules (in mathematical form) can be found in the appendix \ref{appendix:complete-set-of-rule} to this thesis.

\subsubsection{Melodic Rules}

Fux explains that there are rules that apply within parts (the horizontal rules) about the order of the interval between one note and the next: we find, for example, that a melody is more beautiful\footnote{Throughout this work we will speak of the "beauty of music". This beauty is highly subjective, and therefore reference will be made to the Fuxian concept of music to define whether a melody is beautiful or not. In other words, music will be considered beautiful if it conforms to the rules of Fux, and vice versa} if the intervals between its successive notes are small, that there is no chromatic succession, if the notes that follow each other are varied, and so on. These 'horizontal' rules are called 'melodic' rules because they concern only the melody. These rules apply within a given counterpoint.


\subsubsection{Harmonic Rules}


If there is a horizontal perspective to counterpoint, there is also, of course, a vertical perspective. This perspective is expressed in a harmonic relationship between the different voices. At each point in the composition, a series of rules (known as 'harmonic rules' because they concern harmony alone) apply. For example, there is the rule that in the first beats of each measure the interval between the voices must be a consonance \footnote{A third, a fifth, a sixth or an octave.}; imperfect consonances\footnote{Thirds and sixths.} are preferred to fifths, which are preferred to octaves; and the rule that the different voices must be different at each point in the composition. These rules apply between the counterpoints.

\subsubsection{Motion rules}
Finally, there is a third type of rule: movement rules. These rules are a hybrid of the two discussed above in that they consider not only vertical interaction, i.e. harmony, but also horizontal interaction, i.e. melody. They can therefore be seen as 'diagonal' rules that relate the unique melody of each counterpoint to its respective harmonies. These rules include, for example, the fact that we prefer voices that move in opposite directions (i.e. if one voice goes up, we want the other to go down), the rule that there should be no sequence of fifths or octaves between voices, or the rule that we should not arrive at a perfect interval by direct motion.

\subsubsection{Preferences}\label{subsection:preferences-vs-hard-rules}
This last point does not really concern a kind of rule as such, but rather the preferences that Fux expresses in his work. These preferences are advice that Fux offers in order to write a nice counterpoint. As their name suggests, preferences are optional and not compulsory to follow, as other rules would be.

Fux is never clear about whether a rule\footnote{We use the generic term "rule" to refer to both mandatory rules and preferences.} is a preference or a strict rule --- and that's normal, what he conveys is mostly intuition, and human beings are quite capable of understanding whether a rule is a preference or an obligation; Fux probably didn't expect someone to try to formalise his work three centuries later.

These preferences should be respected whenever possible, and if not, so be it. Here's a good example: Fux indicates that we prefer to have as many different notes as possible in the composition. This is not an absolute rule, but a preference. The more variety there is in the composition, the more beautiful it will be, and the more preferable it will be.

\subsection{Species counterpoint}\label{section:species-counterpoint}
When we talk about species counterpoint, we're talking about five categories of counterpoint. Each species is a separate concept, each with its own peculiarities. We'll go into more detail about the different species below. Let's first concentrate on how species counterpoint works. When composing counterpoint, the starting point is a fixed melodic line, the \cf, which is a basic melody composed entirely of whole notes. It is the basis of a composition when writing counterpoint. It is from this voice and in relation to it that the others are composed. It is important to note that once the composition is complete, the \cfs is no more or less important than the other voices, and has the same melodic independence as the other voices. It is therefore nothing more than the basis from which we begin to write.

Let's take a look at the five species:
\begin{enumerate}
    \item \textbf{First species}: Note against note \textendash{} the counterpoint is composed entirely of whole notes, and the composition is a sequence of harmonies sounding on the first beat between the counterpoint and the other voices.
    \item \textbf{Second species}: Half notes against whole notes \textendash{} the counterpoint is composed entirely of half notes, which introduce dissonant harmonies.
    \item \textbf{Third species}: Quarters against whole notes \textendash{} the counterpoint is made up entirely of quarter notes, which allow more different movements and more freedom in the composition.
    \item \textbf{Fourth species}: The ligature \textendash{} the counterpoint is delayed by two beats, creating syncopation. The notes are all round (i.e., tied minims, since they span between two measures).
    \item \textbf{Fifth species}: Florid counterpoint \textendash{} counterpoint is a mixture of all the other species and is the richest form of counterpoint. It allows great freedom of composition while respecting the rules of the other types.
\end{enumerate}

Although these types could technically be combined to form a three-part composition with two different types of counterpoint, Fux seems to prefer to write a 'special' counterpoint (i.e. second, third, fourth or fifth species) in one voice and only whole notes (i.e. a first species counterpoint) in another voice. He also sometimes says that it is possible and recommended to mix species, but does not do so extensively.

\subsection{Building a computer tool for writing species counterpoint}
We've just explored the longstanding tradition of counterpoint, a musical style shaped by countless generations of composers. As technology advanced, the idea of automating counterpoint composition emerged. An early attempt, by Schottstaedt in 1984 \cite{bill1984}, involved an expert system based on Fux's rules. His approach used over 300 if-else clauses, but this method had obvious limitations compared to what modern constraints are capable of.

Indeed, it's crucial to understand that if-else clauses are unidirectional, whereas constraints are bidirectional. Schottstaedt's algorithm relied heavily on specific conditions, captured by numerous if-else clauses. In contrast, modern constraint systems support bidirectional constraints with multiple arguments. These systems don't just find single solutions, they represent sets of potential solutions. This flexibility is a significant improvement over the directional nature of if-else clauses.

Furthermore, constraint systems offer an advantage in specifying intricate search heuristics. This adaptability and efficiency highlight the stark contrast between the outdated approach of if-else clauses and the modern capabilities of bidirectional constraint systems in the realm of counterpoint composition.

\paragraph{}
In 1997, a genetic programming and symbiosis approach to automatic counterpoint generation was developed. A team from Michigan used this approach to optimise counterpoints of the 5th species and make them more attractive \cite{polito1997musica}. A similar approach was used in 2004 to generate fugues (hence counterpoint), also using genetic algorithms \cite{garay2004fugue}. The results are quite promising, and generate more than interesting results, but the end result is still far from being able to provide a complete counterpoint composition.

Many years later, in 2010, a team of researchers from the University of Malaga developed an automated method for the generation of first-species counterpoint using probabilistic logic \cite{Aguilera2010}. Their approach was specifically tailored to compositions in C major, providing a generated counterpoint in response to a given \cf. Please note that this application only evaluates the harmonic attributes of the counterpoint, ignoring the melodic aspect.

Two years later, a team from London developed a way to generate high-quality first-species counterpoint using a variable neighbourhood search algorithm \cite{Herremans2012}. Their research was limited to first-species counterpoint, but they addressed issues such as preferences (finding the best counterpoint) and user-friendly interface. Once again, their results are more than impressive, but their research is limited to the first two-voice species.

Finally, a research was carried out in 2015 on Fux's counterpoint \cite{komosinski2015automatic}, with the aim of generating the first species counterpoint using dominance relations, has yielded fairly good results. The search demonstrates the use of this paradigm and its applicability, and is a good starting point for composing counterpoints of other species based on the same concept.

\paragraph{}
If we now focus on applications that have gone as far as the user interface and are now ready to be used, we should mention two namesakes, both called 'Counterpointer', which have the merit of offering a functional tool for composing counterpoint.

The first Counterpointer tool \cite{counterpointer_ms}, which anyone\footnote{Anyone... or almost, as it is a paid tool.} can use to check the validity (or not) of their counterpoint. Its last release was in 2019 as a desktop application, and it works like this: an apprentice composer tries to write a counterpoint, and then submits it to the tool. The tool then decides whether the counterpoint is valid according to the traditional rules of counterpoint\footnote{Not only Fux's rules, but also those of other authors.}. It also provides feedback to help the student composer improve their future counterpoint writing. The tool is not able to write counterpoints automatically, nor is it explicit about how it works, as it is completely closed source and has no accessible report. It is therefore impossible to know the paradigm it uses or the exact rules it follows.

Another attempt at automatic counterpoint writing is the Counterpointer project in 2021, created by a team of students at Brown University as part of a software engineering course \cite{counterpointer_project}. The project is less accomplished than the aforementioned application, but it has the merit of being able to generate two-voice counterpoints of the first, second and third species. It is an entirely free and open source project. While the results are encouraging, the project has been discontinued as it was a course project and their method of finding a counterpoint seems much less efficient than the efficiency that a constraint solver can achieve. 


\paragraph{}
This brief overview leads us to conclude that there is no satisfactory tool for composing counterpoint in a user-friendly way, with good quality, quickly and with several voices. It is to fill this gap that this research has been carried out. This was the aim of T. Wafflard's thesis and it is therefore natural that this thesis should follow in his footsteps.



\section{Standing on the shoulders of giants: underlying works and editions of \gaps used}
As has just been said, this work is the continuation of T. Wafflard's work. However, it also relies heavily on the work of
\begin{itemize}
    \item \textcite{GiLthesis}, who presented an interface for using Gecode functions in Lisp called "GiL". This interface was then tested with some rhythm-oriented constraints.
    \item \textcite{Melothesis}, who explored the use of constraint programming in OpenMusic using GiL. The tool that was produced in this thesis is capable of producing songs with basic harmonic and melodic constraints.
    \item \textcite{Melo2thesis}, who created a tool capable of combining the strengths of the first two implementations while continuing to develop support for GiL.
\end{itemize}

As with T. Wafflard, the musical reference work chosen is Fux's \gap, because it is a pillar of counterpoint theory and because it is fairly easy to extract rules from it (although Fux is sometimes very vague about his intentions). And as with any book published several centuries ago (1725 in the case of \gap), there are many versions and translations. This is good news, as Fux can sometimes be really unclear about what he means, and having many versions (some annotated, some not) from many people who also had to interpret Fux to translate it is a great treasure, as it helps to clarify Fux's meanings. This work is therefore based on several different editions and translations of the book, although it is mainly based on Alfred Mann's English translation \cite{GaPEng}. French (both Chevalier's \cite{GaPFrChevalier} and Denis's \cite{GaPFrDenis}), German \cite{GaPDe} and Latin \cite{GapLa} translations are used when it is necessary to remove an ambiguity or clarify an unclear rule. These translations have been chosen because French is the \textit{lingua franca} of the team; German is the language of Fux and the environment in which he developed; and Latin is the original version, so we can hope that it is the most faithful to what he wanted to convey.

\section{Tools and implementation}\label{subsection:tools-and-implementation}
This subsection discusses the implementation of the automatic counterpoint generator. To do so, we briefly explain how constraint programming works, the tools used by FuxCP, and the actual implementation of it.

\subsection{Constraint Programming}
Constraint Programming (CP) is a programming paradigm used to solve large combinatorial problems, such as planning and scheduling problems. It works by defining constraints between variables that limit the values these variables can potentially take \cite{rossi2008constraint}. In doing so, the domains of these variables are reduced. R. Barták explains in a very clear way what a constraint really is, as he describes in his "Guide to Constraint Programming" \cite{bartak1998constraint}:

\begin{quote}
    A constraint is simply a logical relation among several unknowns (or variables), each taking a value in a given domain. A constraint thus restricts the possible values that variables can take, it represents some partial information about the variables of interest. For instance, "the circle is inside the square" relates two objects without precisely specifying their positions, i.e., their coordinates. Now, one may move the square or the circle and he or she is still able to maintain the relation between these two objects. Also, one may want to add other object, say triangle, and introduce another constraint, say "square is to the left of the triangle". From the user (human) point of view, everything remains absolutely transparent.
\end{quote}

\paragraph{Constraint propagation}
Each time a constraint is defined, the domain of the affected variables is reduced according to the possibilities left by the constraint. This is called constraint propagation. Let's imagine a variable $A$ whose domain is $\{0, 1, 2, 3\}$ and a variable $B$ whose domain is $\{0, 1, 2\}$. When the constraint $A<B$ is applied, the domain of $A$ is reduced to $\{0, 1\}$, because the new constraint makes it impossible for $A$ to have a value of $2$ or $3$ without violating the constraint.

A constraint solver is an implementation that systematically browses the search space in look for a solution. A given problem can have many solutions, just one, or even none, if the domains are too small or the constraints contradict each other, and the system becomes overconstrained.
A solution is found when all variables are fixed, i.e. their domain is reduced to a single value. We know that a problem has no solution if a domain is empty after a constraint has been declared (because this means that no value can be found for that variable that does not violate the constraint).

\paragraph{Branching}
Obviously, declaring constraints is not enough to magically find a solution. In the example we proposed earlier (with A and B), the single constraint placed on the search space doesn't allow us to determine the values of A or B, and their domains remain composed of more than one value. 
To actually find a solution, the constraint performs a branching. That is, it studies two antagonic possibilities and splits the search space into two subproblems accordingly. More specifically, it chooses a variable and studies the case where that variable is equal to a certain value, and the case where it is not equal to that value. For example, the solver might study the case where $B=0$ (the first branch) and the case where $B \neq 0$ (the second branch). If the solver finds an inconsistency in either case, it knows that the entire branch can be discarded (as it does not lead to a valid solution). Immediately after branching, the solver again performs constraint propagation, since constraint propagation occurs whenever any domain is modified, and consists of adjusting all domains to which the modified domain is linked by a constraint. In our case, after setting $B$ to $0$, the solver propagates all constraints linked to $B$, i.e. the only constraint in our problem (i.e. $A<B$). This affects the domain of $A$, reducing it to an empty domain (because no value in the domain of $A$ is less than $0$). The solver, noticing that one domain is empty, concludes that this branch contains no solution and therefore knows that the only possible branch is the other one, i.e. the one in which we assumed $B\neq 0$. We can therefore safely remove $0$ from the domain of $B$, since this value is not contained in any solution. Repeating the branching process each time it is necessary produces three solutions: $A=0$ and $B=1$, $A=0$ and $B=2$, and finally $A=1$ and $B=2$.

\paragraph{Heuristics}
As with all problems involving searching a space in quest of a solution, it is very useful to have heuristics allowing for an efficient search. A heuristic is a rule or strategy used to make informed decisions about variable assignments and value choices during the solution search. These rules are designed to exploit the characteristics and structures of the problem to improve the chances of finding solutions more quickly. A common heuristic is variable ordering, where the algorithm selects variables to assign values to based on factors such as the size of the domain or the number of associated constraints. Another important heuristic is value ordering, which determines the order in which values are tested for a given variable assignment. By incorporating heuristics, constraint solvers can prioritise the most promising branches of the search tree, effectively reducing the search space and speeding up the identification of feasible solutions. While heuristics speed up the solving process, it's important to strike a balance between exploration and exploitation, as overly aggressive heuristics risk missing potentially valuable solution paths.


To return to our previous example, a value-ordering heuristic might be "branch first on low values of $A$ and high values of $B$, since we know that we are looking for a solution where $A$ is less than $B$, and we can reason that there are more chances of satisfying this constraint when $B$ is large and $A$ is small.

\paragraph{Costs in Constraint Programming}
While the basis of Constraint Programming is this succession of propagations and branchings, there also exists a notion of cost, which distinguishes between two valid solutions (i.e. satisfying all constraints) into a less good solution and a better solution.


For example, in our simple example, we could define cost as the sum of $A$ and $B$ and want to minimise the cost. This means that we are looking for a valid solution where $A$ and $B$ are as small as possible. In this case, only the first of the three solutions mentioned above is chosen, i.e. $A=0$ and $B=1$, as it is the best possible solution (with a cost of 1).

\paragraph{Branch and Bound}
Branch and Bound is a systematic algorithm used in Constraint Programming to efficiently explore the solution space and find an optimal solution with respect to a given cost or objective function. This technique extends the basic Constraint Programming approach by introducing a mechanism to prune unpromising branches of the search tree, thereby reducing the computational effort required to find the optimal solution. The algorithm starts with the initial problem and iteratively divides it into subproblems, called branches, by making decisions about variable assignments. For each branch, the algorithm evaluates its feasibility and its potential to lead to a better solution. If a branch is deemed infeasible or cannot possibly improve on the current best-known solution, it is pruned from further consideration. This process continues until all branches have been explored, or until the algorithm converges on the optimal solution. Branch and Bound is particularly valuable for large combinatorial problems because it efficiently narrows the search space, allowing the solver to focus on promising regions and accelerating the discovery of optimal solutions in constrained programming scenarios \cite{morrison2016branch}. 

\subsubsection{Relevance of the use of constraint programming in the field of music}
The question now is: what is the connection between this problem-solving method and counterpoint composition? Well, it turns out that music is actually a very good application of constraint programming: you can represent all aspects of a composition by variables, set rules between these variables, and the solver takes care of finding a valid solution. In fact, in music, it's never just one factor that determines whether the music is beautiful, but an interaction of many. And as humans, it's sometimes difficult to find a valid solution (i.e. to compose music that sounds good) because the range of possibilities and the interactions between factors are so numerous.  However, as we have just seen, exploring a search space in which a large number of constraints are defined is something that a constraint solver does very well. The most arduous task then becomes putting the rules that make music beautiful down on paper, and this is the task that many musicologists and composers have set themselves. Once these rules have been defined by musicologists and composers, it is "simply" a matter of formalising them and passing them to the constraint solver so that it can compose a melody that respects these rules.

What's more, by defining a rigorous way of distinguishing a good composition from a bad one, the solver can even find increasingly beautiful solutions.

\subsection{Open Music}
OpenMusic is a powerful and innovative visual programming environment, written in CommonLisp, designed specifically for composers, researchers and musicians involved in computer-aided composition \cite{OpenMusic}. Developed by the Institute for Research and Coordination in Acoustics/Music (IRCAM) in Paris, OpenMusic provides a graphical interface that allows users to create and manipulate musical structures using a variety of predefined modules. This visual programming language facilitates the representation of complex musical ideas, algorithms and data flows through a user-friendly interface, making it accessible to both novice and experienced composers. 


FuxCP is a library for OpenMusic, which means that OpenMusic is the interface for using FuxCP. Any user wishing to use FuxCP writes their \cfs (FuxCP's input) into OpenMusic and then launches the solution search from within OpenMusic. Specifically, FuxCP retrieves the \cfs from OpenMusic and then defines the constraint problem. When a solution is found, it is passed to OpenMusic and the user gets it as an OpenMusic object.

\subsection{Gecode and GiL}
After receiving the input (which is the \cf), FuxCP uses Gecode to define the search space and starts searching for a solution. Gecode, short for Generic Constraint Development Environment, is an open source toolkit for developing constraint-based systems \cite{Gecode}. It provides a high-level C++ library for efficiently modelling and solving constraint problems. Gecode supports a wide range of constraints, variables, and search strategies, making it a versatile platform for tackling combinatorial problems such as scheduling, optimisation, and configuration. 


Since FuxCP is an OpenMusic library written in Lisp, it is naturally written in Lisp too. To be able to use the Gecode C++ library, FuxCP uses the Gecode Interface Lisp (GiL), which is a wrapper that allows Gecode functions to be used in Lisp programs. It in turn relies on The Common Foreign Function Interface (CFFI), which is an interface for calling C++ functions in Lisp \cite{CFFI}.  GiL is far from complete, but it provides enough tools to solve interesting constraint problems in Common Lisp.

\subsection{Concrete implementation}
To put it all together: FuxCP gets its input (the \cf) from the user interface, which is OpenMusic. It then defines a constraint programming problem in Gecode, using GiL. 

As for the way it defines the problem, here is a little clarification: first, when the \cfs is received, a whole series of constants and variables are defined: for example, the length of the \cf, the arrays representing the pitches of the counterpoints, ...
Then all these variables are constrained according to the constraints defined in the formalisation of Fux's rules (discussed in chapter \ref{chapter:species}). The constraints are set sequentially: first the constraints on the \cf, then the constraints on the first counterpoint, and finally the constraints on the second counterpoint. A diagram of the code architecture and the integration of FuxCP with other tools can be found in Appendix \ref{fig:softwarearchitecure}.


\section{T. Wafflard's thesis in a nutshell}\label{section:thomas-in-a-nutshell}

In 2023, T. Wafflard proposed a complete formalisation of Fux's two-voice counterpoint \cite{wafflard2023}. This formalisation takes each of the rules given by Fux about two-voice composition and translates them into formal logic. Those formal relations are then translated into constraints and given to a constraint solver. When given an input \cf, the solver applies all the constraints it was given and yields an output counterpoint. The following subsection is not intended to be an exhaustive summary of all of T. Wafflard's excellent work, but rather a brief outline of the idea behind it and the procedure followed. This brief overview is given to the reader because many of the concepts in T. Wafflard's work are at the heart of the three-voice generalisation and will be key to understanding the three-voice formalisation.

\subsection{Variables} \label{Wafflard-variables}
In order to formalise Fux's rules, it was necessary to define variables whose purpose is to represent a compositional reality. Thus T. Wafflard created variables to represent various concepts, such as the pitch of notes (written $N$), the harmonic interval between voices (written $H$), the melodic interval from one note to the next (written $M$), and many others. These variables are then related to each other according to the formalised rules. The constraint solver searches for all possible values of these variables, according to the constraints, and stops when all variables in $N$ (the pitches) are fixed, as this means that a solution has been reached (the notes of the counterpoint are known, and this is the goal of the solver).

Useful constants were also defined and will be reused throughout this work. The most important of these are $m$, which represents the length of the \cfs (and thus the number of bars in the composition), $n$, which represents the number of notes in a given counterpoint, or Cons$_{\text{all, p, imp}}$, which represents the set of consonances (all, perfect and imperfect).

Of course, there are many different solutions for the same \cf, and in order to distinguish between two valid solutions (i.e. all solutions that respect all constraints), some costs have been defined. These costs are intended to convey the preferences expressed by Fux in \gap. In fact, Fux sometimes expresses a preference that is not an absolute rule: "this should be done if possible, but is not necessary". Thus, the solver considers a valid solution with a low cost to be better than a valid solution with a high cost. The costs scale from 0 (when we don't care) to 64$m$ (when something should only happen as a last resort), but most costs scale from 0 to 8. These costs are then added together to form a total cost, which the solver tries to minimise.

\subsection{Array notation}
As we have just seen it is necessary to refer to numerous arrays, that formally represent the musical composition and many of its underlying aspects (pitches, harmony, melody, ...). These arrays always have two indexes: the first index represents the time in question, the second index represents the measure in question. These indices are written in computer notation. For example, $X[3, 7]$ represents the variable $X$ on the 3rd beat of the 7th measure.

Let's put it all together. Here is a simple example of how the variables and constants are brought together through the formalisation: 
\begin{equation}    
    H[0, m] \in Cons_{\text{perf}}
\end{equation}
 This is a rule which means that the last harmonic interval must be a perfect consonance\footnote{Please refer to section \ref{subsection:modified_variables} for more exact details}. All rules from T. Wafflard's thesis can be found in appendix \ref{appendix:complete-set-of-rule}.

 \subsection{In practice}
 In practice, to solve this constraint programming problem, the constraints are written in Lisp, and thanks to the Gecode Interface Lisp (GiL) \cite{GiL}, it uses the Gecode constraint solver \cite{Gecode} to find a solution. To make the constraints work, we need a starting point. The starting point is the \cf (see \ref{section:species-counterpoint}).
 When given a \cf, the solver defines a set of variables (those mentioned in the formalisation) to which constraints are then applied (the relations from the formalisation) and produces a counterpoint that obeys all the rules that have been defined and whose quality can be given by the cost. As explained in \ref{Wafflard-variables}, in the case of T. Wafflard's implementation, the total cost is the sum of all the costs, and this cost is minimised by a depth first search algorithm that finds the lowest cost, and then gives the corresponding solution.
 
 
 As for the front-end, all the user sees and interacts with is OpenMusic \cite{OpenMusic}, an object-oriented visual programming environment for musical composition based on Common Lisp \cite{commonlisp} developed by the Parisian institute IRCAM (Institute for Acoustic/Music Research and Coordination) \cite{IRCAM}.
 
 The present work has exactly the same basic functioning (as is explained in the previous subsection \ref{subsection:tools-and-implementation}). 


\section{The contributions of this thesis}
The aim of this work is to generalise T. Wafflard's formalisation to three-voice counterpoint, still based on Fux's work, and to create the corresponding implementation. It would be too easy to believe (wrongly) that three-voice counterpoint is nothing more than the combination of two two-voice counterpoints. From this point of view, we would then calculate a first counterpoint according to the \cf, and then a second counterpoint again according to the \cf, and that's it. Obviously, this view is too simplistic and doesn't really capture all the interactions between three voices. It is to this point (the peculiarities brought about by the addition of a third voice) that a whole chapter of this thesis is devoted. Another chapter is devoted to translating the rules from \gap into formal logic. A final but not less important section discusses and analysing the impact of costs, and the musicality of the solutions. The following is a more detailed summary of the contributions of this thesis. 
\begin{itemize}
    \item \textbf{Introducing new concepts and redefining variables}:
    As we have just mentioned, a three-part composition is much more than a (two+one) part composition. So we had to redefine and define a whole series of concepts to adapt to this reality. The creation of the (lowest, middle and highest) stratum concept is part of this, and is essential for formalising Fux's counterpoint constraints. All of this is discussed in Chapter \ref{chapter:defining-some-concepts-and-redifing-the-variables}.
    \item \textbf{Mathematical formalisation of three-part counterpoint}: As with the two parts of the formalisation, we rewrote Fux's explanations into unambiguous English and then translated them into logical notation. This formalisation builds on the previous formalisation for two voices, and sometimes (rarely) has to modify it. This formalisation can be found in Chapter \ref{chapter:species}.
    \item \textbf{Implementation of a working constraint solver for a three-voice composition}: Those logical rules were then implemented as constraints and the solver was adapted to allow a search for two counterpoints. The whole code of this implementation can be found in Appendix \ref{chapter:whole-code}, and its architecture in the Appendix \ref{chapter:architecture}.
    \item \textbf{Researching the best way to express Fux's preferences}: Three-part composition introduces so many possibilities for result composition that it is important to rethink the way we think about preferences. These preferences are understood by the solvers as costs (where a preferred solution in Fux's sense has a lower cost to the solver). Therefore, some techniques for managing these preferences are discussed to find out the best way to implement them as costs. This is very important as it allows the solver to produce solutions with high musicality. These techniques are discussed in Chapter \ref{chapter:search}
    \item \textbf{Musical analysis of the solutions generated by the solver}: Finding the best solution also means being able to assess the quality of current solutions. For this, see Chapter \ref{chapter:musicality}. 
    \item \textbf{Adapted user interface to allow them to compose with three voices and set a cost order}: All the new capabilities of the solver and the costing techniques must also be accessible to the user: it is now possible for a user to freely combine any number of species to form a three-part composition, and to set a cost order to indicate their preferences to the solver (in addition to the previous ability to set personalised costs). A guide for this can be found in Appendix \ref{chapter:user-guide}.
\end{itemize}
% citer Fux page 71 