\chapter{Complete set of rules for two and three part compositions}\label{appendix:complete-set-of-rule}
This appendix contains all the rules for composing counterpoint for two or three voices. All the rules apply to three-voice compositions, but only the rules for two voices apply to two-voice compositions. Rules for three-part compositions are indicated by '3V' at the beginning of the rule.


Some of the rules are different depending on whether the composition is for two or three voices. In these cases, the mathematical relationship to be followed for the rule in question, depending on the situation, is clearly indicated.
\section*{Constraints of the First Species}
\subsection*{Harmonic Constraints of the First Species}
\begin{enumerate}[wide, label=\bfseries 1.H\arabic*]
  \item\label{rule:allcons}{\textit{All harmonic intervals must be consonances.}} 
\begin{equation}
    \begin{gathered}
        \forall j \in [0, m)\quad 
        H[0, j] \in Cons
    \end{gathered}
\end{equation}

\item\label{rule:firstpcons}{\textit{The first harmonic interval must be a perfect consonance.}}
When dealing with two-part composition:
\begin{equation}
    \begin{gathered}
        H[0, 0] \in Cons_{p}
    \end{gathered}
\end{equation}

\item\label{rule:lastpcons}{\textit{The last harmonic intervals must be a perfect consonance.}}
When dealing with three-part composition:
\begin{equation}
  \begin{gathered}
      H[0, m-1] \in Cons_{p}
  \end{gathered}
\end{equation}

\item\label{rule:keytone}{\textit{The key tone is tuned according to the first note of the \cfdot}}

\begin{equation}
    \begin{gathered}
        \lnot IsCfB[0, 0] \implies H[0, 0] = 0\\
        \lnot IsCfB[0, m-1] \implies H[0, m-1] = 0
    \end{gathered}
\end{equation}

\item{\textit{The counterpoint and the \cfs cannot play the same note at the same time except in the first and last measure.}}

\begin{equation}
    \begin{gathered}
        \forall j \in [1, m-1)\quad
        Cp[0, j] \neq Cf[j]
    \end{gathered}
\end{equation}

\item{\textit{Imperfect consonances are preferred to perfect consonances.}}


\begin{equation}
    \begin{gathered}
        \forall j \in [0, m)\\
        Pcons_{costs}[j] = \begin{cases}
            cost_{Pcons} & \text{if } H[0, j] \in Cons_{p}\\
            0 & \text{otherwise}
        \end{cases}\\
        \text{moreover } \C = \C \cup \sum _{c \in Pcons_{costs}} c
    \end{gathered}
\end{equation}

\item{and \textbf{1.H8} \textit{The harmonic interval of the penultimate note must be a major sixth or a minor third depending on the \cfs pitch.}}\label{rule:penult-interval-2v}
\addtocounter{enumi}{1} 
\begin{equation}
    \begin{gathered}
        % \lnot IsCfB[0, m-2] \implies H[0, m-2] = 9
        \rho := \max (positions(m)) - 1\\
        H[\rho] = \begin{cases}
            9 & \text{if } IsCfB[\rho]\\
            3 & \text{otherwise}
        \end{cases}\\
        \text{where } \rho \text{ represents the penultimate index of any counterpoint.}
    \end{gathered}
\end{equation}

\subsection*{Melodic Constraints of the First Species}
\end{enumerate}

\begin{enumerate}[wide, label=\bfseries 1.M\arabic*]
\item\label{rule:notritone}{\textit{Tritone melodic intervals are forbidden.} }

\begin{equation}
    \begin{gathered}
        \forpm\\
        M[\rho] = 6 \implies Mdeg_{costs}[\rho] = cost_{tritoneMdeg}\\
    \end{gathered}
\end{equation}

\item\label{rule:mlesixth}{\textit{Melodic intervals cannot exceed a minor sixth interval.}}

\begin{equation}
    \begin{gathered}
        \forj\quad
        M[0, j] \leq 8
    \end{gathered}
\end{equation}

\subsection*{Motion Constraints of the First Species}
\end{enumerate}
\begin{enumerate}[wide, label=\bfseries 1.P\arabic*]

\item\label{rule:nopconsbydm}{ \textit{Perfect consonances cannot be reached by direct motion.}}

When dealing with two-part composition:
\begin{equation}
    \begin{gathered}
        \forj\quad
        H[0, j+1] \in Cons_{p} \implies P[0, j] \neq 2
    \end{gathered}
\end{equation}

When dealing with three-part composition:
\begin{equation} \begin{aligned}
  &\forall j \in [0, m-2) :\\
  &P[0, j] = 2 \land H[0, j+1] \in Cons_{p} \\
  &\iff cost_{\text{{direct\_move\_to\_p\_cons}}}[j] = 8
\end{aligned} \end{equation}

\item\label{rule:codmotions} {\textit{Contrary motions are preferred to oblique motions which are preferred to direct motions.}}

\begin{multicols}{3}
    \begin{itemize}
        \item $cost_{con}$\\ \dft{no cost}
        \item $cost_{obl}$\\ \dft{low cost}
        \item $cost_{dir}$\\ \dft{medium cost}
    \end{itemize}
\end{multicols}

\begin{equation}
    \begin{gathered}
        \forj\\
        P_{costs}[j] = \begin{cases}
            cost_{con} & \text{if } P[0, j] = 0\\
            cost_{obl} & \text{if } P[0, j] = 1\\
            cost_{dir} & \text{if } P[0, j] = 2
        \end{cases}\\
        \text{moreover } \C = \C \cup \sum _{c \in P_{costs}} c
    \end{gathered}
\end{equation}

\item\label{rule:battuta}{ \textit{At the start of any measure, an octave cannot be reached by the lower voice going up and the upper voice going down more than a third skip.}}


\begin{equation}
    \begin{gathered}
        i := \max (\B), \forj\\
        H[0, j+1] = 0 \land P[i, j] = 0 \land \begin{cases}
            M_{brut}[i, j] < -4 \land IsCfB[i, j] \iff \bot\\
            M_{cf}[i, j] < -4 \land \lnot IsCfB[i, j] \iff \bot
        \end{cases}\\
        \text{where } i \text{ stands for the last beat index in a measure.}
    \end{gathered}
    \label{eq:battuta}
\end{equation}
\end{enumerate}

\section*{Constraints of the Second Species}
\subsection*{Harmonic Constraints of the Second Species}
\begin{enumerate}[wide, label=\bfseries 2.H\arabic*]
\item\label{rule:consthesis}{ \textit{Thesis harmonies cannot be dissonant.}}

As explained above, there is no constraint to add because it would be a duplicate of rule \ref{rule:allcons}.

\item\label{rule:arsisdim}{\textit{Arsis harmonies cannot be dissonant except if there is a diminution.}}

\begin{equation}
    \begin{gathered}
        \forj\\
        IsDim[j] = \begin{cases}
            \top & \text{if } M^2[0, j] \in \{3, 4\} \land M^1[0, j] \in \{1, 2\} \land M^1[2, j] \in \{1, 2\}\\
            \bot & \text{otherwise}
        \end{cases}
    \end{gathered}
\end{equation}

\begin{equation}
    \begin{gathered}
        \forj \quad
        \lnot IsCons[2, j] \implies IsDim[j]
    \end{gathered}
\end{equation}

\item\label{rule:penult2nd} \label{rule:penultexception}{and \textbf{2.H4} \textit{In the penultimate measure the harmonic interval of perfect fifth must be used for the thesis note if possible. Otherwise, a sixth interval should be used instead.}}
\addtocounter{enumi}{1}

\begin{equation}
    \begin{gathered}
        H[0, m-2] \in \{7, 8, 9\}\\
        \therefore penulthesis_{cost} = \begin{cases}
            cost_{penulthesis} & \text{if } H[0, m-2] \neq 7\\
            0 & \text{otherwise}
        \end{cases}\\
        \text{moreover } \C = \C \cup penulthesis_{cost}
    \end{gathered}
\end{equation}

\end{enumerate}
\subsection*{Melodic Constraints of the Second Species}
\begin{enumerate}[wide, label=\bfseries 2.M\arabic*]

\item\label{rule:octaveleap}{ \textit{If the two voices are getting so close that there is no contrary motion possible without crossing each other, then the melodic interval of the counterpoint can be an octave leap.}}

\begin{equation}
    \begin{gathered}
        \forj, \forall M_{cf}[j] \neq 0\\
        M[0, j] = 12 \implies (H_{abs}[0, j] \leq 4) \land (IsCfB[j] \iff M_{cf}[j]>0)
    \end{gathered}
\end{equation}

\item\label{rule:notsamecons}{ \textit{Two consecutive notes cannot be the same.}}
When dealing with two-part composition:
\begin{equation}
    \begin{gathered}
        \forp \quad
        Cp[\rho] \neq Cp[\rho+1]
    \end{gathered}
\end{equation}

When dealing with three-part composition:
\begin{equation}
  \begin{aligned}
      &\forall j \in [1, m-1), \quad j \neq m-2:\\
      &((N[2, j-1] \neq N[0, j]) \land (N[0, j] \neq \land N[2, j])) \\
      &\land \\
      & ((N[2, m-3] \neq N[0, m-2]) \lor (N[0, m-2] \neq N[2, m-2]) )
  \end{aligned}
\end{equation}


\end{enumerate}
\subsection*{Motion Constraints of the Second Species}
\begin{enumerate}[wide, label=\bfseries 2.P\arabic*]

\item\label{rule:motion2nd}{\textit{If the melodic interval of the counterpoint between the thesis and the arsis is larger than a third, then the motion is perceived based on the arsis note.}}


\begin{equation}
    \begin{gathered}
        \forj \quad
        P_{real}[j] = \begin{cases}
            P[2, j] & \text{if } M[0, j] > 4\\
            P[0, j] & \text{otherwise}
        \end{cases}
    \end{gathered}
\end{equation}

\item\label{rule:battuta2}{ \textit{Rule \ref{rule:battuta} on the battuta octave is adapted such that it focuses on the motion from the note in arsis.}}
\item 
This constraint already had an adapted mathematical notation in the chapter of
the first species. Note that this constraint would indeed use P[2] and not P$_{real}$.


\end{enumerate}
%%%%%%%%%%%%%

\section*{Constraints of the Third Species}
\subsection*{Harmonic Constraints of the Third Species}
\begin{enumerate}[wide, label=\bfseries 3.H\arabic*]
  \item\label{rule:fivequarters}{ \textit{If five notes follow each other by joint degrees in the same direction, then the harmonic interval of the third note must be consonant.}}

\begin{equation}
    \begin{gathered}
        \forj\\
        % \{M[0, j]\land M[1, j]\land M[2, j]\land M[3, j]\} \leq 2\ \land\\
        % \left(
        %     \{M_{brut}[0, j]\land M_{brut}[1, j]\land M_{brut}[2, j]\land M_{brut}[3, j]\} > 0\ \lor \right. \\
        %     \left.
        %     \{M_{brut}[0, j]\land M_{brut}[1, j]\land M_{brut}[2, j]\land M_{brut}[3, j]\} < 0\
        % \right)\\
        % \implies IsCons[2, j]
        \left(
            \bigwedge_{i=0}^{3} M[i, j] \leq 2
        \right)
        \land
        \left(
            \bigwedge_{i=0}^{3} M_{brut}[i, j] > 0
            \lor
            \bigwedge_{i=0}^{3} M_{brut}[i, j] < 0
        \right)\\
        \implies IsCons[2, j]
    \end{gathered}
\end{equation}


\item\label{rule:thirddiss} {\textit{If the third harmonic interval of a measure is dissonant then the second and the fourth interval must be consonant and the third note must be a diminution.}}


\begin{equation}
    \begin{gathered}
        \forj\\
        IsCons[2, j] \lor \left( IsCons[1, j] \land IsCons[3, j] \land IsDim[j]\right)\\
        \text{where } IsDim[j]=\top \text{ when the \nth{3} note of the measure } j \text{ is a diminution.}
    \end{gathered}
\end{equation}

\item\label{rule:cambiata} {\textit{It is best to avoid the second and third harmonies of a measure to be consonant with a one-degree melodic interval between them.}}


\begin{equation}
    \begin{gathered}
        \forj\\
        Cambiata_{costs}[j] = \begin{cases}
            cost_{Cambiata} & \text{if } IsCons[1, j] \land IsCons[2, j] \land M[1, j] \leq 2\\
            0 & \text{otherwise}
        \end{cases}
    \end{gathered}
\end{equation}

\item\label{rule:penult3sp} {\textit{In the penultimate measure, if the \cfs is in the upper part, then the harmonic interval of the first note should be a minor third.}}

\begin{equation}
    \begin{gathered}
        \lnot IsCfB[m-2] \implies H[0, m-2] = 3
    \end{gathered}
\end{equation}
\end{enumerate}
\subsection*{Melodic Constraints of the Third Species}
\begin{enumerate}[wide, label=\bfseries 3.M\arabic*]
  \item\label{rule:twobeats} {\textit{Each note and its two beats further peer are preferred to be different.}}


\begin{equation}
    \begin{gathered}
        \forpmm \\
        MtwoSame_{costs}[i, j] = \begin{cases}
            cost_{MtwobSame} & \text{if } M^2[\rho] = 0\\
            0 & \text{otherwise}
        \end{cases}
    \end{gathered}
\end{equation}
\end{enumerate}
\subsection*{Motion Constraints of the Third Species}
\begin{enumerate}[wide, label=\bfseries 3.P\arabic*]
  \item\label{rule:motion3rd} {\textit{The motion is perceived based on the fourth note.}}

This implies that the costs of the motions and the first species constraints on the motions are deducted from $P[3]$.
\end{enumerate}
%%%%%%%%%%%%%%%%%%%%%


\section*{Constraints of the Fourth Species}

\subsection*{Motion Constraints of the Fourth Species}
\begin{enumerate}[wide, label=\bfseries 4.P\arabic*]
  \item\label{rule:dissolved} {\textit{Dissonant harmonies must be followed by the next lower consonant harmony.}}

\begin{equation}
    \begin{gathered}
        \forall j \in [1, m-1) \quad
        \lnot IsCons[0, j] \implies M_{brut}[0, j] \in \{-1, -2\}
    \end{gathered}
\end{equation}

\item\label{rule:nosecond} {\textit{If the \cfs is in the lower part then no second harmony can be preceded by a unison/octave harmony.}}

\begin{equation}
    \begin{gathered}
        \forall j \in [1, m-1)\\
        IsCfB[j+1] \implies H[2, j] \neq 0 \land H[0, j+1] \notin \{1, 2\}
    \end{gathered}
\end{equation}

\end{enumerate}
\subsection*{Harmonic Constraints of the Fourth Species}
\begin{enumerate}[wide, label=\bfseries 4.H\arabic*]
  \item\label{rule:arsiscons} {\textit{Arsis harmonies must be consonant.}}

\begin{equation}
    \begin{gathered}
        \forall j \in [0, m-1) \quad
        H[2, j] \in Cons
    \end{gathered}
    \label{eq:arsiscons}
\end{equation}

\item\label{rule:noseventh} {\textit{If the \cfs is in the upper part, then no harmonic seventh interval can occur.}}

\begin{equation}
    \begin{gathered}
        \forall j \in [1, m-1) \quad
        \lnot IsCfB[j] \implies H[0, j] \notin \{10, 11\}
    \end{gathered}
\end{equation}

\item\label{rule:lowpenult4th} \label{rule:uppenult4th} {and \textbf{4.H4} \textit{In the penultimate measure, the harmonic interval of the thesis note must be a major sixth or a minor third depending on the \cfs pitch.}}

\begin{equation}
    \begin{gathered}
        H[0, m-2] = \begin{cases}
            9 & \text{if } IsCfB[m-2]\\
            3 & \text{otherwise}
        \end{cases}
    \end{gathered}
\end{equation}
\end{enumerate}
\subsection*{Melodic Constraints of the Fourth Species}
\begin{enumerate}[wide, label=\bfseries 4.M\arabic*]
  \item\label{rule:fullsyncopations} {\textit{Arsis half notes should be the same as their next halves in thesis.}}


\begin{equation}
    \begin{gathered}
        \forall j \in [0, m-1) \quad
        NoSync_{costs} = \begin{cases}
            cost_{NoSync} & \text{if } M[2, j] \neq 0\\
            0 & \text{otherwise}
        \end{cases}
    \end{gathered}
\end{equation}

\item\label{rule:m2same} {\textit{Each arsis note and its two measures further peer are preferred to be different.}}


\begin{equation}
    \begin{gathered}
        \forall j \in [0, m-1)\\
        MtwomSame_{costs} = \begin{cases}
            cost_{MtwomSame} & \text{if } Cp[2, j] = Cp[2, j+2]\\
            0 & \text{otherwise}
        \end{cases}
    \end{gathered}
\end{equation}

\end{enumerate}
%%%%%%%%%%%%%%